---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, we load the data set:

```{r}
activity <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
We can use tapply and the sum() function to generate daily totals, from which
we can create a histogram:

```{r}
dailySteps <-tapply(activity$steps, activity$date, sum, na.rm = TRUE)
dailySteps
hist(dailySteps, main = "Histogram of Daily Step Totals", xlab = "Daily Steps")
```

We can also take the mean and median of total steps per day:

```{r}
mean(dailySteps)
median(dailySteps)
```


## What is the average daily activity pattern?

What does a "typical day" look like? Let's average the steps by time interval
and plot as a time series:

```{r}
intervalAverages <-tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
plot((names(intervalAverages)), intervalAverages, type = "l", xlab = "Time of Day",
     ylab = "Average Steps")
```

The time of day with the highest average is **8:35 a.m.** :

```{r}
highestAvg <-intervalAverages[order(intervalAverages,decreasing = TRUE)[1]]
highestAvg
```


## Imputing missing values
2304 rows are missing data:

```{r}
nrow(activity[rowSums(is.na(activity))>0,])
```

To replace these, I started by extracting the rows with missing values:

```{r}
naRows <-activity[is.na(activity$steps),]
head(naRows)
```

We're going to replace with the average steps value for that time interval. 

Recall, we already have this in intervalAverages.

I've created a (very narrow usage) function that takes a row number.

The function will:

1. Look up the time interval from column three of that row in naRow

2. Use that value to find the average from intervalAverages

```{r}
getAvg <- function(row) {
     time <- naRows[row,3]
     intervalAverages[as.character(time)]
}
```

We can quickly use sapply to run this over every row in naRows, creating our 
replacement set of the corresponding interval average for each NA value:

```{r}
replaceSet <-sapply(1:nrow(naRows), getAvg)
head(replaceSet)
```

Now to plug that in and fix the original dataset. Since the set we created is in
the same order, it's rather simple:

```{r}
activity[is.na(activity$steps),1] <- replaceSet
```

How does our histogram, mean, and median look with these changes? Well, a lot
more mean-reverting:

```{r}
dailySteps <-tapply(activity$steps, activity$date, sum)
dailySteps
hist(dailySteps, main = "Histogram of Daily Step Totals", xlab = "Daily Steps")
mean(dailySteps)
median(dailySteps)
```


## Are there differences in activity patterns between weekdays and weekends?

We create a factor variable with two levels -- weekday and weekend

```{r}
dayOfWeek <- sapply(as.Date(activity$date),weekdays)
dayOfWeek[dayOfWeek %in% c("Monday","Tuesday","Wednesday","Thursday","Friday")] <- "Weekday"
dayOfWeek[dayOfWeek %in% c("Saturday","Sunday")] <- "Weekend"
weekDayOrEnd <- as.factor(dayOfWeek)
activity <-cbind(activity, weekDayOrEnd)
head(activity)
```

Next, we'll separate the weekend and weekday data:

```{r}
weekendData <- activity[weekDayOrEnd=="Weekend",]
weekdayData <- activity[weekDayOrEnd=="Weekday",]
```

With our data separated, we can aggregate. We'll do weekday first.

I create a data frame of the aggregate data, and add a column that signifies it to 
be a weekday (this will come in handy later!)

```{r}
stepsWeekday <- aggregate(weekdayData$steps,list(weekdayData$interval),mean, na.rm =TRUE)
weekdayLabel <- rep("weekday",times = nrow(stepsWeekday))
stepsWeekday <- cbind(stepsWeekday, weekdayLabel)
head(stepsWeekday)
```

The process is identical for the weekend data:

``` {r}
stepsWeekend <- aggregate(weekendData$steps,list(weekendData$interval),mean, na.rm =TRUE)
weekendLabel <- rep("weekend",times = nrow(stepsWeekend))
stepsWeekend <- cbind(stepsWeekend, weekendLabel)
head(stepsWeekend)
```

With both made, we'll give them synced up variable names so we can rbind:

```{r}
names(stepsWeekend) <- c("Interval","Steps","day_or_end")
names(stepsWeekday) <- c("Interval","Steps","day_or_end")
weeksData <- rbind(stepsWeekday, stepsWeekend)
```

Finally, we need to make the variable *day_or_end* a factor so that we can panel. 
Then we can plot using ggplot2! 

```{r}
weeksData$day_or_end <- as.factor(weeksData$day_or_end)
library(ggplot2)
ggWeek <- ggplot(weeksData, aes(x = Interval, y = Steps)) + geom_line() + 
          ylab("Average Number of Steps") + xlab("Interval") + 
          ggtitle("Averages Steps by Interval: Weekends Vs Weekdays") + 
          facet_grid(weeksData$day_or_end)
ggWeek